---
title: "Google Analytics Capstone - Case Study 2"
author: "Jacob Akinyemi"
date: "06/11/2023"
output:
  html_document:
    df_print: paged
---

# Case Study 2 - How Can a Wellness Technology Company Play It Smart? 

Hello everyone, this notebook illustrates my approach to Bellabeat data analysis case study on Data Analytics by Google Coursera 

I approached this case study by first understanding the business questions that needs to be answered, then i explored the data to find insights and trends. The tools i used in completing this case study includes **RStudio**, **Google Cloud Platform**, and **Tableau**. 

Some of the things i learned after completing this the Bellabeat case study includes: How to ask the right questions and explore data to explore data to make business decisions, also learned about different tools that can be used for a case study and how they can be used. 

To complete this case study, i will go through every phase of data analysis, which is 
* Ask
* Prepare
* Process
* Analyze
* Share
* Act


### Scenario 
 
In this case study, I am playing the role of a junior data analyst working on the marketing analyst team for a company called Bellabeat, a high-tech manufacturer of health-focused product for women. Bellabeat is a successful small company, but they have the potential to become a larger player in the global smart device market. Urška Sršen (aliased as U.S, the co founder & Chief Creative Officer) believes that analyzing smart device fitness data could help unlock new growth opportunities for the company.

I have been asked to focus on one of Bellabeat’s products and analyze smart device data to gain insight into how consumers are using their smart devices which will help guide the marketing strategy for the company.


## Products

* Bellabeat app: Provides users with health data related to their activity, sleep, stress, menstrual cycle, and mindfulness habits.

* Leaf: A wellness tracker that can be worn as a bracelet, necklace, or clip. Connects with the Bellabeat app to track activity, sleep, and stress.

* Time: A wellness watch with smart technology to track user activity, sleep, and stress. Connects to the Bellabeat app to provide insight into daily wellness.

* Spring: A water bottle that tracks daily water intake using smart technology to ensure you are adequately hydrated. Connects to the Bellabeat app to track hydration levels.

* Bellabeat membership: A subscription-based membership program that gives users 24/7 access to fully personalized guidance on nutrition, activity, sleep, health, beauty, and mindfulness-based on their lifestyle and goals.



## About the company

Bellabeat is a high-tech company that manufactures health-focused smart products which are beautifully developed by U.S that informs and inspires women around the world regarding their activity, sleep, stress & reproductive health.

Bellabeat has also invested in traditional advertising media, such as radio, out-of-home billboards, print, and television but focuses on digital marketing extensively such as investing in Google Search, maintaining active Facebook & Instagram pages, consistently engaging consumers on Twitter, running video ads on Youtube and displays ads on the Google Display Network to support campaigns around key marketing dates.

U.S knows that an analysis of Bellabeat’s available consumer data would reveal more opportunities for worth. She has asked the marketing analytics team to focus on a Bellabeat product and analyze smart device usage data in order to gain insight into how people are using their smart devices. Using this information, she would like high-level recommendations for how these trends can inform Bellabeat marketing strategy.



# Ask

U.S has asked me to analyze smart device usage data in order to gain insight into how consumers use non-Bellabeat smart devices. She wants me to select one Bellabeat product to apply these insights to my presentation. Since this phase involves asking SMART(specific, measurable, action-oriented, relevant and time-bound) questions and understanding stakeholders expectations. These questions will guide my analysis:

1. What are some trends in smart device usage?

2. How could these trends apply to Bellabeat customers?

3. How could these trends help influence Bellabeat marketing strategy?


## Guiding questions

1. **What is the problem i am trying to solve?**

The main problem is to analyze how consumers are using non-Bellabeat Smart devices and how the corresponding insights can help determine the marketing strategy for Bellabeat products.


2. **How can my insights drive business decisions?**

My insights will help Bellabeat to grow and be a strong competitor in global smart market usage.



## Key tasks and Deliverables

1. **Identify the business task**

Gain insights from non bellabeat smart devices and formulate a good marketing strategy for Bellabeat.

2. **Consider key stakeholders**

* Primary stakeholders: Urška Sršen and Sando Mur
* Secondary Stakeholders: Bellabeat marketing analytics team




# Prepare

The data i will be using is from FitBit Fitness Tracker [here](https://www.kaggle.com/datasets/arashnic/fitbit). A public dataset that is made available through Mobius [here](https://www.kaggle.com/arashnic). It contains personal fitness trackers from thirty Fitbit users who consented to the submission of personal tracker data, which includes a minute-level output for:

* Physical activity
* Heart rate
* Sleep monitoring

It also includes information about daily activity, steps, and heart rate that can be used to explore user habits.

Since the files are already grouped in a folder, there’s no need to organize them. The names of the files are also fairly easy to recognize given the context of the data, so we will not be modifying them as well.


**ROCCC Analysis**

* Reliability: **LOW** — dataset was collected from 30 individuals whose gender is unknown.

* Originality: **LOW** — third-party data collected using Amazon Mechanical Turk.

* Comprehensive: **MEDIUM** — dataset contains multiple fields on daily activity intensity, calories used, daily steps taken, daily sleep time, and weight record.

* Current: **MEDIUM** — data is 5 years old but the habit of how people live does not change over a few years.

* Cited: **HIGH** — data collector and the source is well documented.



# Process

I am using RStudio for this case study, I am more proficient in RStudio and its packages. 

* Checking the data for errors
* Transforming the data so I can work with it effectively

Lets load install and load the necessary packages required for this process which would be: 

* **Tidyverse**

* **Janitor**

* **Lubridate**

* **Skimr**

## Setting up my environment by installing and loading these packages

## Install packages

```{r}
install.packages("tidyverse")
install.packages("janitor")
install.packages("lubridate")
install.packages("skimr")
```

## Load packages

```{r}
library(tidyverse)
library(janitor)
library(lubridate)
library(skimr)
```

### After this, I will import the datasets into RStudio using **read.csv()**, then i will create two dataframes for the two datasets.


```{r}
#df_name <- read.csv(dataset_location)
daily_activity <- read.csv("dailyActivity_merged.csv")
daily_sleep <- read.csv("sleepDay_merged.csv")
```

## Exploring a few key tables

Lets take a look at the daily_activity data.
```{r}
head(daily_activity)
```

Lets Identify all the columns in the daily_activity data.
```{r}
colnames(daily_activity)
```

Lets take a look at the daily_sleep data.
```{r}
head(daily_sleep)
```


Lets identify all the columns in the daily_sleep data.
```{r}
colnames(daily_sleep)
```
Note that both datasets have the 'Id' field - this can be used to merge the datasets.


Lets inspect our data to see if there are any errors with formatting by using **"str()"**

```{r}
#str(dataframe_name)
str(daily_activity)
```

```{r}
str(daily_sleep)
```

After a brief view of the output, there are few issues that we need to address:

* The naming of the column names (camelCase)

* daily_activity$ActivityDate is formatted as CHR not as a date format

* daily_sleep$SleepDay is formatted as CHR not as a date format



To clean the column names to make sure they are unique and consistent, we would use "clean_names()"

```{r}
daily_activity <- clean_names(daily_activity)
daily_sleep <- clean_names(daily_sleep)
```


Lets format 'daily_activity$ActivityDate' and 'daily_sleep$SleepDay' into proper date format using **"as.date()"**.

```{r}
# Convert string into date using as.date()
daily_activity$activity_date <- as.Date(daily_activity$activity_date, '%m/%d/%y')
daily_sleep$sleep_day <- as.Date(daily_sleep$sleep_day, '%m/%d/%y')
```


After a quick look at our current data, let’s add a day of the week, sedentary hours, and total active hours column for further analysis in daily_activity. I will not be adding a month column since the dataset only provides information collected within a month. 

Let’s also add new columns which convert the current minutes of collection to hours and round it using **round()** in sleep_day. I will also be adding a column to indicate the time taken to fall asleep in sleep_day as well.

```{r}
# Round basically rounds off a number, the syntax would be = round(object, digits = x)
# Select(-c(column_name)) removes a column from a dataframe


daily_activity$day_of_week <- wday(daily_activity$activity_date, label = T, abbr = T)
daily_activity$total_active_hours = round((daily_activity$very_active_minutes + daily_activity$fairly_active_minutes + daily_activity$lightly_active_minutes)/60, digits = 2)
daily_activity$sedentary_hours = round((daily_activity$sedentary_minutes)/60, digits = 2)


daily_sleep$hours_in_bed = round((daily_sleep$total_time_in_bed)/60, digits = 2)
daily_sleep$hours_asleep = round((daily_sleep$total_minutes_asleep)/60, digits = 2)
daily_sleep$time_taken_to_sleep = (daily_sleep$total_time_in_bed - daily_sleep$total_minutes_asleep)
```


## **Remove Outliers**

Before we move onto the phase where we actually start to analyze the data frame, we need to remove any **outliers** from the data.

In this case, let’s remove rows in which the total_active_hours & calories burned are 0. The reasoning behind this is that we’re using data collected from Fitbits, which are wearables. If they don’t wear their smart devices it doesn’t collect information, hence we will remove the clutter from the data frame. Users might have also disabled GPS/accelerometer functions that allow for the collection of steps taken.

```{r}
#In laymans term, '!' means is not equals to
daily_activity_cleaned <- daily_activity[!(daily_activity$calories<=0),]
daily_activity_cleaned <- daily_activity_cleaned[!(daily_activity_cleaned$total_active_hours<=0.00),]
```



# Analyze (Using RStudio)

I will be using ggplot for this section of the analysis phase.

Let’s revisit our business task to ensure we are not plotting or trying to hypothesize information/relationships which will not help in solving the business tasks which are:

1. What are some trends in smart device usage?

2. How could these trends apply to Bellabeat customers?

3. How could these trends help influence Bellabeat's marketing strategy?


After having a brief view of the current data, I will be plotting a few observations revolving around:

1. *The average*: Steps taken, sedentary hours, very active minutes & total hours asleep.

2. Which days are users the most active?

3. The relationship between total active hours, total steps taken, and sedentary hours against calories burned.



Let’s have a quick look at the average steps taken, sedentary hours, very active minutes & total hours of sleep using **“summary()”**.

```{r}
summary(daily_activity_cleaned$total_steps)
```
```{r}
summary(daily_activity_cleaned$sedentary_hours)

```

```{r}
summary(daily_activity_cleaned$very_active_minutes)

```
```{r}
summary(daily_sleep$hours_asleep)

```
With a brief view of the outputs above:

* The average number of steps per day was 8319, which is within the 6000–8000 recommended steps per day, however, 25% of people do not hit that recommended quota.

* The average sedentary hours were 15.87 hours, which is absurdly high, shattering the recommended limit of **7–10 hours**.

* The average very active minutes also falls short of the recommended 30 minutes of vigorous exercise every day. Only 25% of people manage to hit this quota.

* The average hours spent asleep (6.9) also barely hits the quota of the recommended sleep time of 7–9 hours.



Let’s have a look at which days are users most active:

```{r}
# options(scipen=) will remove any scientific notations

options(scipen = 999)
ggplot(data = daily_activity_cleaned) +
  aes(x = day_of_week, y = total_steps) +
  geom_col(fill =  'red') +
  labs(x = 'Day of week', y = 'Total steps', title = 'Totap steps taken in a week')
ggsave('total_steps.png')

```



```{r}
ggplot(data = daily_activity_cleaned) +
  aes(x = day_of_week, y = very_active_minutes) +
  geom_col(fill =  'green') +
  labs(x = 'Day of week', y = 'Total very active minutes', title = 'Total activity in a week')
ggsave('total_activity.png')
```



```{r}
ggplot(data = daily_activity_cleaned) +
  aes(x = day_of_week, y = calories) +
  geom_col(fill =  'blue') +
  labs(x = 'Day of week', y = 'Calories burned', title = 'Total calories burned in a week')
ggsave('total_calories.png')
```


As we can see, the most active days for Fitbit users were Sundays, with a slow decline throughout the week. This could be due to motivation levels being fairly high during the end of the week.



Next, let’s investigate the relationship between total active hours, total steps taken, and sedentary hours against calories burned by using the following:

```{r}
# We use geom_smooth() to aid in seeing pattern in the presence of overplotting (when all the plots are too scattered and or too closed)

ggplot(data = daily_activity_cleaned) +
  aes(x= total_active_hours, y = calories) +
  geom_point(color = 'red') +
  geom_smooth() +
  labs(x = 'Total active hours', y = 'Calories burned', title = 'Calories burned vs active hours')
ggsave('calories_burned_vs_active_hours.png')


```


```{r}
ggplot(data = daily_activity_cleaned) +
  aes(x= total_steps, y = calories) +
  geom_point(color = 'orange') +
  geom_smooth() +
  labs(x = 'Total steps', y = 'Calories burned', title = 'Calories burned vs total steps')
ggsave('calories_burned_vs_total_steps.png')

```


```{r}
ggplot(data = daily_activity_cleaned) +
  aes(x= sedentary_hours, y = calories) +
  geom_point(color = 'purple') +
  geom_smooth() +
  labs(x = 'Sedentary hours', y = 'Calories burned', title = 'Calories burned vs sedentary hours')
ggsave('sedentary_hours_vs_calories_burned.png')
```


At a glance, we can tell that there is a positive correlation between calories burned and total steps taken/total active hours. However, in the last chart, we can see that the correlation is confusing.

I was expecting an inverse relationship with the first 2 charts however I was wrong. The relationship between sedentary hours and calories burned was fairly positive up till about the 17-hour mark.






# **Share and Act**


In the previous section of Analyze & Share, we have covered the 1st and 2nd business task which are:


1. What are some trends in smart device usage

2. How could these trends apply to Bellabeat customers (I believe that displaying the trends would already indicate how Bellabeat customers would follow suit). 


Based on my findings after my analysis, I would like to share my hypothesis on this. 
Users spend more time engaged in physical activity specifically on Sundays, which then proceeds to wane throughout the week with a slight peak on Thursdays which then sees a slow climb on Saturdays.
I suspect that:

(Motivation levels & free time are higher on the weekends, which would provide an opportunity for users to sneak in a workout.
As work load decreases, a window of opportunity to exercise would present itself in the midweek (Thursdays)
We see an alltime low of recorded activity on Friday’s due to the possibility of social engagement with friends/coworkers after working hours)


Now to answer the final business task, I would like to share my recommendations based on my findings to help influence Bellabeat’s marketing strategy.

1. Bellabeat could host events limited to those that are enrolled in their Bellabeat memberships which would reward users who engage in a healthy lifestyle(IE 8k steps a day, less than 7 hours sedentary, etc.) with points. With enough points, users could then use points to purchase products that help supplement a healthy lifestyle.

2. Bellabeat could partner with brands (IE wellness, sports, health) to reward users who consistently engage in a healthy lifestyle with coupons/store discounts.

3. With the 2 previous points combined, Bellabeat could select previously unhealthy individuals (who are now healthy), interview them and publish motivational videos as to how Bellabeat encouraged them to have a change in lifestyle.


Next, I would provide some general recommendations to further improve Bellabeat’s products:


1. Bellabeat could implement personalized milestones, to encourage users to slowly engage in a more healthy lifestyle. A simple way of doing this is to create some sort of AI companion on the app/product that would be grumpy/sad if the user does not hit the milestone.


2. Bellabeat could implement a simple reminder to inform users that they’ve been sedentary for too long by indefinitely vibrating the device until the device picks up movement/increase in heart rate, which would indicate that they’ve engaged in some sort of physical activity.


**Additional Remarks**

* Bellabeat should require users to input their height and their activity levels so that BMR calculations and a more accurate calculation of TDEE would be possible. This would aid future analysis as well.

* Bellabeat should create devices that would track sleep more sophisticatedly (IE REM sleep tracking, deep sleep tracking) to provide more insights into sleep health, as in the dataset provided, we only had the quantity of sleep, not the quality of sleep.
